---
title: "Practice Lecture 2 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "Jan 30, 2024"
---

## Matrices

Now that we finished vectors, let's learn about the matrix object. Let's use the dimension of the learning problem in class.

```{r}
n = 100 #number of historical objects: the people
p = 3 #number of features about each

random_junk = round(runif(n * p), 2)
random_junk
X = matrix(random_junk, nrow = n, ncol = p)
X
class(X)
X[1, 1] #the square brackets are still the getter but since there's two dimensions there's two indices
X[1, ] #first row
class(X[1, ]) #note: not matrix anymore
X[1, , drop = FALSE]
class(X[1, , drop = FALSE]) #still matrix... this matters sometimes and we will see an example of it
X[, 1] #first column
class(X[, 1]) #note: not matrix anymore
X[, 1, drop = FALSE]
class(X[, 1, drop = FALSE])
#other ways of subsetting matrices
X[1 : 4, ]
X[, 2 : 3]
X[, c(1, 3)]
X[seq(1, n, by = 2), ]
X[1, 1] = 1984 #the squarae brackets are still the setter
X[1, 1] = "1984" #warning: this will wind up casting all values in the matrix to string
head(X)
```

Sometimes handy but not recommended is getting / setting with only one index.

```{r}
head(X)
X[1] #first entry i.e. the 1,1 entry
X[2] #first entry i.e. the 2,1 entry (column major indexing, see https://cran.r-project.org/web/packages/reticulate/vignettes/arrays.html)
X[n * p] #last entry i.e. n,p entry
X[n * p + 1] #beyond last entry returns missing / illegal
X[-1] #returns everything but the first entry and it automatically vectorizes
X[1 : 100] #returns the first 100 i.e. the first column
X[X > 0.5] = NA #replace all values no matter what row / col they exist in
X
```


A few handy functions for numerical matrices:

```{r}
X = matrix(random_junk, nrow = n, ncol = p) #reset
rowSums(X)
colSums(X)
rowMeans(X)
colMeans(X)
```

## Conditional statements and looping

We will now learn how to program in R.

# Conditional Statements (and ways to print to screen too).

```{r}
x = 3
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else {
  print("The value of x is"); print(x) #each gets its own line
}

if (x > 0){
  ##
} 
else { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!! This is not Java or C++!
  ##
}

x = -5
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else if (x < -3) { ##NOTE: ELSE IF MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!!
  print(paste("The value of x is", x, collapse = " ")) #paste with collapse is like a string join
} else {
  print("The value of x is"); print(x) #each gets its own line
}
```

There is a one line convenience function called ifelse:

```{r}
x = -10 : 10
b = -10 : 10
a = ifelse(x > 0, b, -1)
a
x = -5 : 5
a = ifelse(x > 0, 1, -1)
a
```

There is also a switch in R. It is _not_ an official keyword (despite RStudio's endorsement), but it is useful for non-vector character expressions:

```{r}
ch = "two"
# ch = c("one", "two")
switch(ch,
   "one" = 1,
   "two" = 2,
   "three" = 3,
   -99
)
```


# Looping

A for loop contains three ingredients: (a) a looping variable joined with keyword "in" to (b) a vector or list (which we will discuss later) and executes over (c) an expression recommended to be enclosed in curly braces:

```{r}
for (i in 1 : 10){
  my_loop_variable = i
  print(my_loop_variable)
}
```

A slight difference from most other functional languages: after the loop is executed both the looping variable (i) still exists and the variable(s) within the scope of the looped expression (my_loop_variable) still exist:

```{r}
i
my_loop_variable
```

It is recommended to "clean up your namespace" whenever you can.

```{r}
for (i in 1 : 10){
  my_loop_variable = i
  print(my_loop_variable)
}
rm(i, my_loop_variable)
```

Remember, these two variables would vanish in C++ / Java when the for loop ended as it goes out of scope.

You can iterate over a vector as well Python-style or Java-generic-style:

```{r}
vec = seq(1, 10)
for (i in vec){
  print(i)
}
```

And the vector does not have to be numeric:

```{r}
vec = c("a", "b", "c")
for (ch in vec){
  print(ch)
}
```

There is also the while loop but not a "do-while" loop

```{r}
x = 1
while (x <= 10){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
}
```

This can also be done using the while-true and a break:

```{r}
x = 1
while (TRUE){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The "repeat" keyword is the same as the while-true. It is slightly more esoteric but still good to know:

```{r}
x = 1
repeat {
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The final control keyword tells the loop to break the current iteration but continue to the next:

```{r}
xs = rep(NA, 10)
xs[3] = 3
xs[7] = 7
xs

tot = 0
for (x in xs){
  if (is.na(x)){
    next #move to next element in the list in the for loop
  }
  tot = tot + x
}
tot
```

Note: "real" R programmers avoid loops since they are slow. Instead they use the "apply" function (but not "sapply" or "lapply" which are not faster). We will cover these in the future. The latest trend is to use in-line C++ code (we will get to this).

## More data types

Before we get back to modeling, it is worth knowing a couple more data structures in R. These are not "data [science] types", these are "[computer science] data types" that are used frequently in data science applications. 

The first are "lists" which are "ordered hashmaps" or "ordered dictionaries" or "hash tables". if You don't know what this is, you should read about this online as it should have been covered in a intro to CS class.

```{r}
dict = list()
dict$a = "first"
dict$b = "second"
dict$c = "third"
dict
length(dict)
names(dict) #keys
dict_unlisted = unlist(dict) #values
dict_unlisted
class(dict_unlisted) #i.e. a vector
#three ways to access values by key / ordered location
dict$a
dict[["a"]]
dict[[1]] #value of first entered key
class(dict[[1]])
dict[[1 : 2]] #bombs
#now let's try to access a value for a non-existent key / ordered location
dict$q
dict[["q"]]
dict[[4]] #bombs
#convenient means to subset the list
dict[1]
class(dict[1])
dict[1 : 2]
dict[1 : 4] #this is the reason this type of access is not recommended
dict = list("first", "second", "third") #no key => value... what happens?
dict #default keys are the numbers 1, 2, ...
dict[[1]]
dict = list("a" = "first", "b" = "second", "c" = "third") #key => value
dict
dict = list(a = "first", b = "second", c = "third") #key => value
dict
```

There is no "hash set" native to the language. You can create a dictionary and fill it with missings:


```{r}
dict = list()
dict$a = NA
dict$b = NA
dict$c = NA
dict
length(dict)
names(dict) #keys
```

Lists conveniently allow all sorts of data types (as values only).

```{r}
varied_dict = list()
varied_dict$a = "first"
varied_dict$b = 2
varied_dict$c = 1 : 7
varied_dict$d = matrix(NA, nrow = 2, ncol = 2)
varied_dict[["some function"]] = function(x){x^2} #this key is not recommended
varied_dict
varied_dict$`some function` #note the tick marks (sometimes seen) needed due to spaces in key name
length(varied_dict)
names(varied_dict)
```

They have lots of uses in data science applications. We will likely see them in class and if not, you'll definitely see them in the real world. Note that data.frame objects are implemented as lists as well as many other common R objects.

Unfortunately, list can only truly accept characters as keys. If you really need more flexibility here, we will need a library (coming soon).


We will now discuss arrays i.e. multidimensional vectors

```{r}
x = array(1 : 5, 5)
x
X = array(1 : 25, dim = c(5, 5))
X
X = array(1 : 125, dim = c(5, 5, 5))
X
X[1, , ]
X[, 1, ]
X[, , 1]
X[1, 1, 1]
```

These can be associative arrays too and operate like a hash of vectors across arbitrary dimensions:

```{r}
X = array(1 : 125, 
          dim = c(5, 5, 5),
          dimnames = list(
            c("A", "B", "C", "D", "E"),
            c("I", "II", "III", "IV", "V"),
            c("blue", "red", "green", "yellow", "orange")
          ))
X
X["A", , ]
X[, "III", ]
X[, , "orange"]
X["C", , "orange"]
X["C", "IV", "orange"]
```


* Functions

```{r}
my_function = function(x){
  x
}
##You may be used to:
# int my_function(int x){
#  //something
#  return x;
# }
```

* Functions are objects in R. This is why you are actually assigning the function to an object.
* We don't need to declare type since R is not "statically typed" (higher level languages usually are not statically typed). Objects can be coerced into different types on the fly (R is "dynamically typed").
* No need for a "return" statement as the last line is the data that is returned. It is considered bad style to use "return" in R.

Let's make sure this works:

```{r}
my_function(3)
my_function("asd")
my_function(x = 3) #you can specify that 3 is the value for argument "x"
my_function(y = 3) #illegal argument
(function(x){x + 1})(3) #anonymous function or "lambda" (see https://en.wikipedia.org/wiki/Anonymous_function)
```

R is somewhat user friendly as it allows for default argument values, making those arguments optional to specify when calling the function:

```{r}
my_function = function(x = 1, y = 2, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #default execution
my_function(p = 0) #one optional argument specified, others defaulted
my_function(y = -2, q = 0) #two optional arguments specified, others defaulted

my_function = function(x = 1, y, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #no dice
my_function(1, 0) #required argument specified
my_function(y = 0, q = 7) #required argument specified and one optional argument
rm(my_function) #can be deleted since it's an object
```

There are also common functional programming functions. 

* Reduce uses a binary function to successively combine the elements of a given vector and a possibly given initial value. 
* Filter extracts the elements of a vector for which a predicate (logical) function gives true. 
* Find and Position give the first or last such element and its position in the vector, respectively. 
* Map applies a function to the corresponding elements of given vectors. 

If you like this, there are many packages that extend this and organize it nicely e.g. `purrr` (we will get to packages next class).

```{r}
x = c(1, 2, 3, 4, 5) * 10
x
Reduce(sum, x)
Filter(function(x){x <= 30}, x)
Find(function(x){x <= 30}, x)
Position(function(x){x <= 30}, x)
Map(function(x){x + 100}, x) #what happened here?? Map will return a list (for flexibility)
unlist(Map(function(x){x + 100}, x)) #collect values into a vector and the keys became the names of the indices
```

## Libraries in R

So far we've only made use of "base R". This is the funcionality included from a vanilla installation of R. 

R has a huge worldwide community of contributors. Some contribute to newer versions of base R, but the vast majority of contributors contribute by creating and maintaining open-source "R packages" or "libraries". I'm one of the latter - check it out [here](https://scholar.rpkg.net/authordetails.php?author=Adam+Kapelner). 

Many libraries come preinstalled. For instance, the MASS library which stands for "Modern Applied Statistics with S" (a famous textbook of R). We can call a function from the MASS library via the following:

```{r}
MASS::as.fractions(0.99)
MASS::as.fractions(pi)
```

But we cannot just execute the following as the function is not included in "base R".

```{r}
as.fractions(pi)
```

We made use of the scope operator "::" to access a namespace beyond the usual "global namespace" which we've been used to. Parenthetically, you can use the ":::" to access the private / internal functions and variables. Anyone who understands object-oriented programming with public interfaces / APIs would cringe at this!!!

If we are using the MASS library a lot, using the scope operator may get annoying. So similar to the "with" command, we can call

```{r}
library(MASS)
```

which loads all public methods (aka "exported" functions) into the public namespace. 

Now, after the library invocation we can do the following and treat it as a normal function:

```{r}
as.fractions(pi)
```

Is this always a good idea? No... everytime you call `library` it "dirties" the namespace by putting all the functions there and rewriting over functions there previously. This is bad because you are more likely to get namespace conflicts. For instance. Let's say package `kapelner` had a weird `sample` function. This would be clear:

```{r}
v = rnorm(100)
kapelner::sample(v)
sample(v)
```

The first line is doing the special sample function and the second is using base R's sample. But if I do this:

```{r}
library(kapelner)
#...
#...
###10,000 lines of code in which you forget about the fact that the kapelner library is loaded
#...
#...
sample(v)
```

You may think you're getting base R sample function, but you're not and now you're in bug-city! You would have to do the following to be explicit:

```{r}
library(kapelner)
sample(v)
base::sample(v)
```

This is not a recommended thing to do. It's also not recommended for package developers to name functions the same as common base R functions. But this doesn't stop them!

Back to real packages... the content for the MASS package was sitting on the hard drive since it comes with R. But what if you want to use a package that does not come with R? We'll have to install the package just like pip for Python, Rubygems for Ruby, R has a package management system built in. For example, here's a useful package for time series / finance stuff:

```{r}
install.packages("tseries")
```

Note that it knew where to go online - it went to a CRAN mirror. CRAN is the official repository for R packages. Now that it's installed, step 2 is to load it into namespace so we can more seamlessly access its functionality.

```{r}
library(tseries)
```

That was a welcome message.

This library is really cool e.g.

```{r}
ibm_stock_history = get.hist.quote(instrument = "IBM", start = "2018-01-01", end = "2018-02-01")
ibm_stock_history
```

Is this a data frame?

```{r}
class(ibm_stock_history)
```

Nope - they made their own data type. If we had a unit on "writing your own R packages", I would explain how this is done but alas there is no time...

Let's say you're sharing your code with someone and one of your lines is loading a library e.g.

```{r}
library(a_library_my_computer_does_not_have_installed_yet)
```

And my computer doesn't have this library. Then we need to stop what we're doing and install. This could be annoying. Here is a convenience: use the pacman package that installs if necessary:

```{r}
if (!require("pacman")){install.packages("pacman")} #installs pacman if necessary but does not load it!
pacman::p_load(devtools) #ensures that devtools gets installed and loaded into the workspace but pacman does not (very tidy)!
```

It is typical to then have a few lines declaring all packages on top of your R/Rmd script file. Here is an example header from one of my projects. 

```{r}
#if (!require("pacman")){install.packages("pacman")}
#pacman::p_load(knitr, randomForest, dplyr, tidyverse, doParallel, xtable, pracma, yaml)
```

We will be seeing this in pretty much all our demos in the future. It is very rare to be coding in R without making use of packages beyond base R. I'm going to require the use of pacman for HW / projects, etc. It just makes code easier to share, run, etc.

The devtools package is important for modern R usage. It allows downloading R packages directly from source that are not even on CRAN. This allows you to get "bleeding edge" features. For example:

```{r}
install_github("yihui/knitr")
```

However this doesn't always work!

```{r}
install_github("hadley/ggplot2")
```

Why can this fail? Because the computer you're running this on is not setup for compiling C++. Admittedly, MAC's usually succeed here and Windows usually fails here. To make it succeed you need to install a separate program beyond R called Rtools. This is one of the big advantages of using Linux and MAC over Windows - Windows just is more buggy when it comes to "real coding" and it gets in the way when you're out there trying to get stuff done. Linux absolutely is the best here and because Linux is usually the production environment anyway, it may make sense to use it for all your assignments and coding anyway.

Note, you can use the pacman library for this type of installation too. So your header becomes:

```{r}
if (!require("pacman")){install.packages("pacman")}
pacman::p_load(devtools)
pacman::p_load_gh("hadley/ggplot2")
```


# Convenient Mapping Function for Lists with the purrr package

We first load the library.

```{r}
pacman::p_load(purrr)
```

We will see later that the library `purrr` is part of a collection of libraries called the `tidyverse`.

Now imagine you have a collection of objects in a list. For example, let's let the object be matrices with different sizes:

```{r}
my_matrix_list = list()
my_matrix_list[["first"]] = matrix(rnorm(9), nrow = 3)
my_matrix_list[["second"]] = matrix(rnorm(12), nrow = 2)
my_matrix_list[["third"]] = matrix(rnorm(8), nrow = 4)
my_matrix_list
```

And you want to operate on each of those objects and return a list. Let's say I want to get back the dimensions, or the first rows, or the average values and return the same keys:

```{r}
my_dims_list = modify(my_matrix_list, ~ dim(.x))
my_dims_list
my_first_rows_list = modify(my_matrix_list, ~ .x[1, ])
my_first_rows_list
my_avgs_list = modify(my_matrix_list, ~ mean(.x))
my_avgs_list
```

This is a very convenient function known as "mapping" in functional programming. It saves a few lines of code e.g. the first `modify` would be:

```{r}
my_dims_list = list() #make new list to store keys --> dimensions of original matrices
for (key in names(my_matrix_list)){ #iterate over all list by key
  .x = my_matrix_list[[key]] #get value at the key for this iteration
  my_dims_list[[key]] = dim(.x) #run function on value and save it to new list
}
my_dims_list
```

The above which takes 5 lines and is repeated again and again and again in code all takes one line using the `modify` function. 

The `modify` function uses funky syntax which is not standard R. And it doesn't have to be; packages are allowed to extend the language and use symbols to create their own little mini-language. The `.x` above is a dummy variable for the value in the iteration in the imagined for loop (like in my rewritten boilerplate code above). The "~" tilde symbol we will be seeing in base R later on in class but in a completely different context. Here it just means "run the following function".

Modify is just one of the functions in the `purrr` package. See the following cheatsheet for more convenient functions: https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf.


## Loading datasets from R packages

Since R is a language built for data and statistics, it has a ton of interesting data sets by default and even more that are contained in packages. There is really just one command to know:

```{r}
rm(list = ls())
data(iris) #load the iris dataset (as a data frame). This dataset is included in the package "datasets" which is autoloaded by default
class(iris)
?iris
#3 things I always do immediately upon getting a dataset
head(iris)
str(iris)
summary(iris)
```

Here is another very famous dataset

```{r}
MASS::Boston #this just references the object but does not load it into the environment
data(Boston) #error since package MASS is not loaded by default
data(Boston, package = "MASS") #package argument not needed if package loaded 
head(Boston)
```

Most data sets are names some descriptive name like "loandata" or "cars". R has so many datasets. Here they all are by package installed:

```{r}
data(package = .packages(all.available = TRUE))
```


## Errors and Warnings

You can write better functions if you make use of errors and warnings. Java forces you to catch likely errors via the "throws" designation for a method but there is no such requirement in R.

* Errors are unrecoverable, they halt execution i.e. red lights
* Warnings (under usual execution) do not halt execution, but they display a message, i.e. yellow lights

Here's how they work:

```{r}
my_vector_sum = function(xs){
  
  if (!(class(xs) %in% c("numeric", "integer"))){ #short for class(xs) == "numeric" | class(xs) == "integer"
    stop("You need to pass in a vector of numbers not a vector of type \"", class(xs), "\".\n") #throw error!
    # warning("Your vector of type \"", class(xs), "\" will be coerced to numbers.\n") #throw warning!
    # xs = as.numeric(as.factor(xs))
  }
  
  tot = 0
  for (x in xs){
    tot = tot + x
  }
  tot
}
my_vector_sum(c(1, 2, 3))
my_vector_sum(c("a", "b", "c"))
```

There is a try-catch as well:

```{r}
xs = c("a", "b", "c")
tot = my_vector_sum(xs)

tot = tryCatch(
  {
    my_vector_sum(xs)
  },
  error = function(e){
    print("xs was non-numeric... coercing xs to numeric...")
    my_vector_sum(as.numeric(as.factor(xs)))
  }
)
tot
```

The recommended thing to do of course is to query if it is non-numeric within the function `my_vector_sum` and cast it then. Possibly create an argument toggling this behavior on/off with a default of off.


## Matrix operations in R

R can do all the standard matrix operations. Let's go through them quickly. First initialize two example matrices:

```{r}
A = matrix(rep(1, 4), nrow = 2)
A
B = array(seq(1, 4), dim = c(2, 2))
B
I = diag(2) #create an identity matrix of size 2x2
I
```

Now we show off some operations:

```{r}
A * B #element-wise multiplication
A %*% B #matrix multiplication
B %*% I
t(B) #transpose
solve(B)
solve(A) #BOOM - why?

#alternative method - to be wrapped in a function
tryCatch(
  {
    solve(A)
  },
  error = function(e){
    print("matrix not invertible, doing Moore-Penrose generalized pseudoinverse instead...")
    MASS::ginv(A)
  }
)

#how would you wrap this and handle it in the real world?
solve(I)
#rank(A) = 1 #no such function... but... there are tons of add-on libraries for matrix computations e.g.
pacman::p_load(Matrix) #load the Matrix library
rankMatrix(B)
rankMatrix(A)
rankMatrix(I)
```

Note that vectors and matrices are not the same:

```{r}
v = c(1, 2, 3) #3-d vector
t(v) #converts to 1x3 vector... unsure why
t(t(v))
v %*% v #seems to default to dot product
t(v) %*% t(t(v)) #dot product
I = diag(3)
I %*% v #seems to default correctly!
I %*% t(v) #actually non-comformable
```

