---
title: "Practice Lectures Week 12 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "May 2, 2022"
---




#Classification Trees and Confusion Tables

Let's load up the adult dataset where the response is 1 if the person makes more than $50K per year and 0 if they make less than $50K per year.

```{r}
pacman::p_load_gh("coatless/ucidata")
data(adult)
adult %<>% 
  na.omit #kill any observations with missingness
```

Let's use samples of 2,000 to run experiments:

```{r}
train_size = 2000
train_indices = sample(1 : nrow(adult), train_size)
adult_train = adult[train_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL
test_indices = sample(setdiff(1 : nrow(adult), train_indices), train_size)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL
```

Make a tree:

```{r}
tree_mod = YARFCART(X_train, y_train)
tree_mod
```

How "big" is this tree model?

```{r}
get_tree_num_nodes_leaves_max_depths(tree_mod)
```

What are the "main" splits?

```{r}
illustrate_trees(tree_mod, max_depth = 4, open_file = TRUE)
```

Compute in-sample and out of sample fits:

```{r}
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Let's look at the confusion table in-sample:

```{r}
table(y_train, y_hat_train)
```

There are no errors here! Thus, precision and recall are both 100%. This makes sense because classification trees overfit.

Let's do the same oos:

```{r}
oos_conf_table = table(y_test, y_hat_test)
oos_conf_table
```

We didn't do as well (of course). Let's calculate some performance metrics. We assume ">50k" is the "positive" category and "<=50k" is the "negative" category. Note that this choice is arbitrary and everything would just be switched if we did it the other way.

```{r}
n = sum(oos_conf_table)
n
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])
acc = (tp + tn) / n
acc
misclassifcation_error = 1 - acc
misclassifcation_error
precision = tp / num_pred_pos
precision
recall = tp / num_pos
recall
false_discovery_rate = 1 - precision
false_discovery_rate
false_omission_rate = fn / num_pred_neg
false_omission_rate
```

Let's see how this works on a dataset whose goal is classification for more than 2 levels. Note: this is only possible now with trees!

```{r}
rm(list = ls())
pacman::p_load(mlbench, skimr)
data(LetterRecognition)
LetterRecognition = na.omit(LetterRecognition) #kill any observations with missingness
skim(LetterRecognition)
?LetterRecognition
```

Now we split the data:

```{r}
test_samp = 500
train_indices = sample(1 : nrow(LetterRecognition), test_samp)
ltr_train = LetterRecognition[train_indices, ]
y_train = ltr_train$lettr
X_train = ltr_train
X_train$lettr = NULL
test_indices = sample(setdiff(1 : nrow(LetterRecognition), train_indices), test_samp)
ltr_test = LetterRecognition[test_indices, ]
y_test = ltr_test$lettr
X_test = ltr_test
X_test$lettr = NULL
```

And fit a tree model and its in-sample and oos fits:

```{r}
tree_mod = YARFCART(X_train, y_train)
y_hat_train = predict(tree_mod, X_train)
y_hat_test = predict(tree_mod, X_test)
```

Take a look at the in-sample confusion matrix:

```{r}
table(y_train, y_hat_train)
```

Perfecto... as expected... 

Now the oos confusion matrix:

```{r}
oos_confusion_table = table(y_test, y_hat_test)
oos_confusion_table
```

Hard to read. Let's make it easier to read by blanking out the diagonal and looking at entried only >= 5:

```{r}
oos_confusion_table[oos_confusion_table < 2] = ""
diag(oos_confusion_table) = "."
oos_confusion_table
mean(y_test != y_hat_test)
```

What's it using to determine letter?

```{r}
illustrate_trees(tree_mod, max_depth = 3, open_file = TRUE)
```

Where did these features comes from?? Deep learning helps to create the features from the raw pixel data. Wish I had a whole next semester to discuss this...

Random Forests:

```{r}
num_trees = 500
train_size = 2000

training_indices = sample(1 : nrow(adult), train_size)
adult_train = adult[training_indices, ]
y_train = adult_train$income
X_train = adult_train
X_train$income = NULL

mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_rf = YARF(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
```

And test:

```{r}
test_indices = sample(setdiff(1 : nrow(adult), training_indices), 25000)
adult_test = adult[test_indices, ]
y_test = adult_test$income
X_test = adult_test
X_test$income = NULL

y_hat_test_bag = predict(mod_bag, X_test)
y_hat_test_rf = predict(mod_rf, X_test)

oos_conf_table_bag = table(y_test, y_hat_test_bag)
oos_conf_table_rf = table(y_test, y_hat_test_rf)
oos_conf_table_bag
oos_conf_table_rf
miscl_err_bag = mean(y_test != y_hat_test_bag)
miscl_err_rf = mean(y_test != y_hat_test_rf)
miscl_err_bag
miscl_err_rf

cat("gain: ", (miscl_err_bag - miscl_err_rf) / miscl_err_bag * 100, "%\n")
```

And on letters:

```{r}
test_samp = 2000
train_indices = sample(1 : nrow(LetterRecognition), test_samp)
ltr_train = LetterRecognition[train_indices, ]
y_train = ltr_train$lettr
X_train = ltr_train
X_train$lettr = NULL
test_indices = sample(setdiff(1 : nrow(LetterRecognition), train_indices), test_samp)
ltr_test = LetterRecognition[test_indices, ]
y_test = ltr_test$lettr
X_test = ltr_test
X_test$lettr = NULL
```

And fit a tree model and its in-sample and oos fits:

```{r}
mod_bag = YARFBAG(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_rf = YARF(X_train, y_train, num_trees = num_trees, calculate_oob_error = FALSE)
mod_bag
mod_rf

y_hat_test_bag = predict(mod_bag, X_test)
y_hat_test_rf = predict(mod_rf, X_test)

oos_conf_table_bag = table(y_test, y_hat_test_bag)
oos_conf_table_rf = table(y_test, y_hat_test_rf)
oos_conf_table_bag
oos_conf_table_rf
miscl_err_bag = mean(y_test != y_hat_test_bag)
miscl_err_rf = mean(y_test != y_hat_test_rf)
miscl_err_bag
miscl_err_rf

cat("gain: ", (miscl_err_bag - miscl_err_rf) / miscl_err_bag * 100, "%\n")
```

Very real gains for classification.



