---
title: "Practice Lectures Week 5 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "Feb 28, 2022"
---


## Assessing quality of a simple linear regression

Regenerate the data from last week:


```{r}
set.seed(1984)
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
h_star = beta_0 + beta_1 * x
epsilons = rnorm(n, mean = 0, sd = 0.33)
y = h_star + epsilons 

pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

Note: our $\mathcal{A}$ was ordinary least squares. What follows below is a method of assessing model fit quality not only for the least squares line, or any linear fit, but any regression fit.

```{r}
simple_df$yhat = b_0 + b_1 * simple_df$x
simple_df$e = y - simple_df$yhat

sse = sum(simple_df$e^2)
mse = sse / (n - 2)
rmse = sqrt(mse)
sse
mse
rmse

s_sq_y = var(y)
s_sq_e = var(simple_df$e)
rsq = (s_sq_y - s_sq_e) / s_sq_y
rsq

#calculated in a different, but equivalent way
sse_0 = (n - 1) * s_sq_y
(sse_0 - sse) / sse_0
```

Let's take a look at $R^2$ visually. We compute null residuals (the $e_0$'s) and model residuals (the $e$'s) and plot a them.

```{r}
simple_df$e_0 = y - mean(y)

ggplot(simple_df) + 
  geom_histogram(aes(x = e), fill = "darkgreen", alpha = 0.3) + 
  geom_histogram(aes(x = e_0, fill = "red", alpha = 0.3)) +
  theme(legend.position = "none")

ggplot(simple_df) + 
  stat_density(aes(x = e), fill = "darkgreen", alpha = 0.3) + 
  stat_density(aes(x = e_0, fill = "red", alpha = 0.3)) +
  theme(legend.position = "none")
```

Note residuals always have sample average = 0 (modulo numeric error):

```{r}
mean(simple_df$e_0)
mean(simple_df$e)
```

We will prove this fact later in class.

Let's take a look at predictions by truth:


```{r}
ggplot(simple_df, aes(x = yhat, y = y)) + 
  geom_point() + 
  xlim(0, max(simple_df$yhat, y)) + 
  ylim(0, max(simple_df$yhat, y)) + 
  xlab("yhat") + 
  coord_fixed() + 
  geom_abline(intercept = 0, slope = 1, color = "orange")
```

Linear regression is pretty popular so there's obviously support for this in R. Before we talk about this, we need to discuss another object type in R. It is called the "formula" object. Here's an example:

```{r}
simple_model_formula = as.formula("y ~ x")
simple_model_formula
```

You can use a convenience:

```{r}
simple_model_formula = y ~ x
simple_model_formula
```

How did this work? R interprets this as a formula because it sees the tilde inside (you have to dig pretty deep into the R language to understand the tilde operator but feel free to ignore it now).

By default the formula object when executed prints out the string you supplied. But obviously it is not just a string. This object contains instructions to model `y` with `x`. This may seem opaque now but you'll see how this works soon.

Getting back to support for the default linear model. The popular function for that implements least squares linear regression is loaded into R automatically in the package `stats`. Here is a list of packages that are loaded into R by default (save the `ggplot2` and RStudio's addition):

```{r}
search()
```

The function `lm` runs least squares. Let's see it:

```{r}
simple_linear_model = lm(simple_model_formula)
simple_linear_model
```

You can skip a step by putting the formula in-line (this is how it's usually done):


```{r}
simple_linear_model = lm(y ~ x)
simple_linear_model
class(simple_linear_model)
```

What did this do? By specifying the formula that $y$ should be modeled with $x$ sing the simple linear model $y = w_0 + w_1 x$
By default it prints out $b_0$ and $b_1$. You can store the vector via:

```{r}
b = coef(simple_linear_model)
b
names(b)
class(b) #i.e. a vector of numbers dimension 2 where each entry is named
```

You can query the linear model about its fit as well:

```{r}
names(summary(simple_linear_model))
summary(simple_linear_model)$r.squared #the R^2
summary(simple_linear_model)$sigma #the RMSE
```

Cleanup...

```{r}
rm(list = ls())
```


## Simple Linear Regression with an example data set

Load up the famous Boston Housing data

```{r}
?MASS::Boston
Xy = MASS::Boston
head(Xy)
```

We would like to see how each feature relates to the response, `medv`. This is a quick and dirty way to do it:

```{r}
for (feature in setdiff(colnames(Xy), "medv")){
  plot(ggplot(Xy, aes(x = Xy[, feature], y = medv)) + geom_point() + xlab(feature))
}
```

Let's try to explain `medv` using the feature `rm` in a simple linear regression (least squares) model.

```{r}
x = Xy$rm
y = Xy$medv

r = cor(x, y)
s_x = sd(x)
s_y = sd(y)
ybar = mean(y)
xbar = mean(x)

b_1 = r * s_y / s_x
b_0 = ybar - b_1 * xbar
b_0
b_1
```

and we can plot this line atop the data:

```{r}
simple_viz_obj = ggplot(Xy, aes(x = rm, y = medv)) + geom_point()
simple_ls_regression_line = geom_abline(intercept = b_0, slope = b_1, color = "red")
simple_viz_obj + simple_ls_regression_line
```

And how well did we do?

```{r}
yhat = b_0 + b_1 * x #this is the g(x^*) function!
e = y - yhat
sse = sum(e^2)
mse = sse / length(y)
rmse = sqrt(mse)
sse
mse
rmse

s_sq_y = var(y)
s_sq_e = var(e)
rsq = (s_sq_y - s_sq_e) / s_sq_y
rsq
```

SSE is not a super useful number alone. MSE is not super useful alone. RMSE is... what does it mean? What does $R^2$ mean?


```{r}
Xy$null_residuals = y - mean(y)
Xy$residuals = e

ggplot(Xy) + 
  stat_density(aes(x = residuals), fill = "darkgreen", alpha = 0.6, adjust = 0.5) + 
  stat_density(aes(x = null_residuals, fill = "red", alpha = 0.6, adjust = 0.5)) +
  theme(legend.position = "none")
```

This is not a great model. Why? Three sources of error... what do you think are the biggest sources of error?

```{r}
rm(list = ls())
```

Let's do this again using R's `lm` function. Here we can leverage the data frame object by using the names of the variables in our model formula:

```{r}
mod = lm(medv ~ rm, data = MASS::Boston)
```

I read this as "build a linear model where we explain median household value using the average number of rooms in the Boston housing dataset". One line! We can of course ping the model for everything we've been talking about:

```{r}
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

And ggplot has amazing integration. Here it runs the model internally and gives smoothed confidence bands (we did not discuss this):

```{r}
ggplot(MASS::Boston, aes(rm, medv)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

```{r}
rm(list = ls())
```


Let's take a look at another dataset.

```{r}
?MASS::Cars93
cars = MASS::Cars93
cars
```

Usually, we are trying to build a model for `Price`. Let's see how `Horsepower` is related to price:

```{r}
pacman::p_load(ggplot2)
ggplot(cars, aes(Horsepower, Price)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```


```{r}
simple_linear_model = lm(Price ~ Horsepower, data = cars)
coef(simple_linear_model)
summary(simple_linear_model)$r.squared
summary(simple_linear_model)$sigma
```

62\% is pretty good $R^2$! But the RMSE is about \$6,000. Using the empirical rule heuristic, that means you can only predict within $\pm \$12,000$ around 95\% of the time. Not so good!

# Predictions with linear models in R

After the model is fit, you may want to predict with it using the $g$ function. Of course R can do this:

```{r}
predict(simple_linear_model, data.frame(Horsepower = 200)) 
#i.e. yhat = g(400)
predict(simple_linear_model, data.frame(Horsepower = c(200, 300, 500))) 
#i.e. the yhat vector = [g(200), g(300), g(500)]
```


##Simple regression with nominal variables

Let's take a look at the factor variable as the predictor. We'll use `Origin` in this example:

```{r}
table(cars$Origin)
```

We can plot how this looks:

```{r}
ggplot(cars, aes(Origin, Price)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

Note that ggplot cannot handle fitting a line with a factor variable.

However...

```{r}
simple_linear_model = lm(Price ~ Origin, data = cars)
coef(simple_linear_model)
summary(simple_linear_model)$r.squared
summary(simple_linear_model)$sigma
```

What happened here? The `lm` function can handle factors. It picks a level to be the reference category (in this case, it's origin = USA) and the fitted slope $b_1$ would be the difference between non-USA and USA. Does it make sense that the slope is positive? Yes - foreign cars charge transport fees and there are a lot of luxury foreign cars.

Why is $R^2$ so low?? Remember the null model is one $\bar{y}$, this model is just two $\bar{y}$'s. How much variance can you possible expect to explain with just two possible prediction values?? Now take a look at RMSE. It's about \$10,000. Before, it was about \$6,000 and the $R^2$ = 62\%. Yes the RMSE has gone up in this case. $R^2$ is not proportion standard error explained, it's proportion variance and that squared error is very different than its square root.

Let's cast this predict as numeric and run it again just to make sure it will be the same. We first code a new dummy variable:

```{r}
cars$origin_is_not_usa = ifelse(cars$Origin == "non-USA", 1, 0)
```

and then we model using this new dummy variable:

```{r}
simple_linear_model = lm(Price ~ origin_is_not_usa, data = cars)
coef(simple_linear_model)
summary(simple_linear_model)$r.squared
summary(simple_linear_model)$sigma
```

Note the reference category is USA and the "non-USA" coefficient indicates the difference in sample averages.

Note that now ggplot can handle the line:

```{r}
ggplot(cars, aes(origin_is_not_usa, Price)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

Let's code the dummy variable differently to take a look at the equivalent regression but this time with the reference category as non-USA

```{r}
cars$origin_is_usa = ifelse(cars$Origin == "USA", 1, 0)
```

```{r}
simple_linear_model = lm(Price ~ origin_is_usa, data = cars)
coef(simple_linear_model)
summary(simple_linear_model)$r.squared
summary(simple_linear_model)$sigma
```

The coefficients here are like "the opposite" in some sense of what we just saw. 

And of course $R^2$ and RMSE are equivalent - it's the same linear model with the same information, just coded differently.

What is the intercept is left out? You can remove the intercept from the formula by adding "0 + ..." which means "no intercept but ..."

```{r}
simple_linear_model = lm(Price ~ 0 + Origin, data = cars)
coef(simple_linear_model)
summary(simple_linear_model)$r.squared
summary(simple_linear_model)$sigma
```

What did $R^2$ seem like it changed?? The `lm` method is calculating $R^2$ differently here if there is no intercept. The null model it is comparing to is $g_0 = 0$ and not our $g_0 = \bar{y}$. This is a small point about the implementation in R and I won't ever test you on this trivia.

Let's do an example of a categorical nominal variable with L>2 levels.

```{r}
summary(cars)
table(cars$Type)
mean(cars[cars$Type == "Compact", "Price"])
mean(cars[cars$Type == "Large",   "Price"])
mean(cars[cars$Type == "Midsize", "Price"])
mean(cars[cars$Type == "Small",   "Price"])
mean(cars[cars$Type == "Sporty",  "Price"])
mean(cars[cars$Type == "Van",     "Price"])
lm(Price ~ 0 + Type, cars)
lm(Price ~ Type, cars)
```

What is the reference level in this variable? Can we change it?

```{r}
levels(cars$Type)
cars$Type = relevel(cars$Type, ref = "Van")
lm(Price ~ Type, cars)
lm(Price ~ 0 + Type, cars)
```


## Correlation and Covariance

Let's load up the Boston Housing data again.

```{r}
boston = MASS::Boston
str(boston)
```

Let us take a look at some covariances and correlations with the response, the median home value.

```{r}
pacman::p_load(ggplot2)
cov(boston$rm, boston$medv)
cor(boston$rm, boston$medv)
ggplot(boston, aes(rm, medv)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
cov(boston$indus, boston$medv)
cor(boston$indus, boston$medv)
ggplot(boston, aes(indus, medv)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

Ever wonder why it's called $R^2$?

```{r}
summary(lm(medv ~ rm, boston))$r.squared
cor(boston$rm, boston$medv)^2
summary(lm(medv ~ indus, boston))$r.squared
cor(boston$indus, boston$medv)^2
```

## Multivariate linear regression

We want to run a multivariate linear regression $\mathcal{H}$ employing the least squares $\mathcal{A}$ manually using our derived linear algebra. Let us first pull out $\mathbb{D}$ as $y$ and $X$.

Let's ensure we augment the `X` to include the 1 vector in front. We need this for the intercept in the $w$ vector in our spec, $\mathcal{H}$.

```{r}
y = MASS::Boston$medv
X = cbind(1, MASS::Boston[, 1 : 13])
head(X)
```

Can we find $X^\top X$?

```{r}
XtX = t(X) %*% X
```

The data frame is great, but unfortunately R does not allow us to use matrix algebra on it.

So let's create a matrix. Note: there are no factor variables with more than one level. `chas` is a binary variable and that's okay. If there were factors with more than level, the following will not work. We will explore this later.

```{r}
X = as.matrix(cbind(1, MASS::Boston[, 1 : 13]))
head(X)
```

So $p = 12$ and $p + 1 = 14$.

Let's make each predictor name nice just for aesthetic value:

```{r}
colnames(X)
colnames(X)[1] = "(intercept)" #this is the standard way lm denotes it (which we will compare to later)
colnames(X)
```


Can we find $X^\top X$?

```{r}
XtX = t(X) %*% X
dim(XtX)
```

Is it full rank?

```{r}
XtXinv = solve(XtX)
```

It worked. This means $X$ is full rank i.e. there is no linear duplication of information over the `13 + 1` predictors. In case we're in doubt:

```{r}
pacman::p_load(Matrix)
rankMatrix(X)[[1]]
rankMatrix(t(X))[[1]]
rankMatrix(XtX)[[1]]
rankMatrix(XtXinv)[[1]]
```


Let's calculate the LS solution then:

```{r}
b = XtXinv %*% t(X) %*% y
b
```

Interpretation: if `crim` "increases" by 1, $\hat{y}$ increases by... etc etc. How would `crim` increase? Big philosophical topic which we are punting on (for now). If all predictors are 0, then $y$ would be predicted to be the intercept, 20.65. Strange concept... not usually important.

What would $g$ look like?

```{r}
g_predict_function = function(x_star){
   x_star %*% b
}
g_predict_function(X[7, ])
y[7] #good prediction!
```

Pretty simple...  and `x_star` could be a matrix of `n_star * (p + 1)` - where `n_star` is however many new observations you wish to predict.

We can compute all predictions:

```{r}
yhat = X %*% b
head(yhat, 10)
```

Can you tell this is projected onto a 14 dimensionsal space from a 506 dimensional space? Not really... but it is...

We can calculate the residuals:

```{r}
e = y - yhat
head(e, 10)
```

What is RMSE?

```{r}
SSE = t(e) %*% e
MSE = 1 / (nrow(X) - ncol(X)) * SSE
RMSE = sqrt(MSE)
SSE
MSE
RMSE
```

Interpret the RMSE...

We can calculate $R^2$ two ways:

```{r}
s_sq_y = var(y)
s_sq_e = var(e)
Rsq = (s_sq_y - s_sq_e) / s_sq_y
Rsq

n = length(e)
SST = (n - 1) * s_sq_y
Rsq = 1 - SSE / SST
Rsq
```

Let's look at distribution of $y$ and $e$ to get an idea about $R^2$ as we did before:


```{r}
pacman::p_load(ggplot2)

ggplot(data.frame(null_residuals = y - mean(y), residuals = e)) + 
  stat_density(aes(x = residuals), fill = "darkgreen", alpha = 0.3) + 
  stat_density(aes(x = null_residuals, fill = "red", alpha = 0.3)) +
  theme(legend.position = "none")
```

What does this tell you about $R^2$?

Now, of course, R has its own function to do all this. We've already seen them! To run a multivariate least squares linear model,

```{r}
mult_lin_mod = lm(medv ~ ., MASS::Boston)
```

No need to (a) create a matrix from the data frame (b) append a 1's column (c) do the linear algebra. It's all done for you. What is this formula `medv ~ .`? Previously we've seen `medv ~ rm` to indicate "fit phenomenon `medv` using predictor `rm`". Here, it's "fit phenomenon `medv` using all available predictors in the data frame". This is a very powerful formula!

Let's extract the estimates $b$ as well as $R^2$ and RMSE:

```{r}
coef(mult_lin_mod)
summary(mult_lin_mod)$r.squared
summary(mult_lin_mod)$sigma
```

Does R offer a simple way to do $g$? Sure...

```{r}
x_star = MASS::Boston[7, ]
y_hat_star = predict(mult_lin_mod, newdata = x_star)
y_hat_star
y[7]
```

If you care about the internals of what R is doing, it retraces our steps perfectly. It first creates the "model matrix" we called the "design matrix" and denoted it X:

```{r}
Xmm = model.matrix(medv ~ ., MASS::Boston)
head(Xmm)
head(X) #same
```

Then it uses an internal function to compute the linear algebra:

```{r}
raw_mod = lm.fit(Xmm, y)
raw_mod$coefficients
```

We will soon see the internals of the `lm.fit` algorithm when we do the linear algebra of QR decomposition.


## OLS using categorical predictors

Note that historically this is called "Analysis of Variance" or "ANOVA" for short. But there is no difference to the computer, it still crunches the same matrices.

Let's get the cars data again:

```{r}
cars = MASS::Cars93
str(cars)
```

Let's try to model `Type`, a factor with 6 levels.

```{r}
table(cars$Type)
```

What will $\hay{y}$ look like? Should be the $\bar{y}$'s for each level. What is $p$? 6. Let' see:

```{r}
anova_mod = lm(Price ~ Type, cars)
coef(anova_mod)
summary(anova_mod)$r.squared
```

The one categorical variable got blown up into 5 features. How to interpret? First need to know the "reference category" i.e. which level is missing in the list. We can see from cross-referencing the coefficient names with the table of the raw feature that the reference category is `Compact`. So what is prediction for the compact type? The intercept. What is prediction of Large type? Intercept + Large, etc. 

What actually happened to get the OLS estimates? Let's see the model matrix:

```{r}
Xmm = model.matrix(Price ~ Type, cars)
head(Xmm, 20)
table(rowSums(Xmm))
```

The predictor `Type` got "dummified" (remember we spoke about this in lecture 1 or 2). There are now 5 dummy variables each representing one of the levels and the reference level is omitted because it is accounted for in the intercept. Let's make sure this is exactly what's going on.

```{r}
y = cars$Price
Xt = t(Xmm) 
XtX = Xt %*% Xmm
XtXinv = solve(XtX)
b = XtXinv %*% Xt %*% y
b
yhat = Xmm %*% b
e = y - yhat
Rsq = (var(y) - var(e)) / var(y)
Rsq
sqrt(sum(e^2) / (nrow(cars) - 6))
```

And of course the coefficients and $R^2$ are identical to the output from `lm`.

If we want to do a more "pure ANOVA", we can get rid of the intercept and see the $\bar{y}$'s immediately. This is handled in R's formula designation by adding a zero:

```{r}
anova_mod = lm(Price ~ 0 + Type, cars)
coef(anova_mod)
```

Is this correct?

```{r}
mean(cars$Price[cars$Type == "Compact"])
mean(cars$Price[cars$Type == "Large"])
mean(cars$Price[cars$Type == "Midsize"])
mean(cars$Price[cars$Type == "Small"])
mean(cars$Price[cars$Type == "Sporty"])
mean(cars$Price[cars$Type == "Van"])
```

What does $R^2$ look like?

```{r}
summary(anova_mod)$r.squared
```

Remember this from last time? What happened? The $R^2$ calculation in `lm` is not accurate without the intercept. Keep this in mind. 

What does the design matrx (model matrix) look like? we can use the `model.matrix` function to generate the columns of $X$ from the data frame. The argument is the formula we wish to generate the model matrix for. Since model matrices don't require

```{r}
Xmm = model.matrix(~ 0 + Type, cars)
head(Xmm, 20)
table(rowSums(Xmm))
```

Very similar. 

Regressions without an intercept are not recommended. Here's why. What if we were doing two factors? I want a linear model with both Type and Airbags:

```{r}
table(cars$AirBags)
```

Airags is another nominal categorical variable, this time with three levels.

We invoke the model as follows.

```{r}
anova_mod = lm(Price ~ Type + AirBags, cars)
coef(anova_mod)
summary(anova_mod)$r.squared
summary(anova_mod)$sigma
```

What are interpretations now? What is the "reference level"? It's actually two levels in one: Type = compact and Airbags = Driver \& Passenger. 

A deeper question: can we read off Type = Midsize and AirBags = none? No... this is a modeling "enhancement" we will discuss in a few lectures from now.

If we model it without an intercept,


```{r}
anova_mod = lm(Price ~ 0 + AirBags + Type, cars)
coef(anova_mod)
```

we only get $\bar{y}$'s for the first factor predictor crossed with the reference category of the second. So above `TypeCompact` refers to the average of Type = Compact and Airbags = Driver \& Passenger.

Now let's create a linear model using one categorical predictor and one continuous predictor. The combination is called for historical reasons "Analysis of Covariance" or "ANCOVA" for short.

Let's use `Type` and `Horsepower`:

```{r}
ancova_mod = lm(Price ~ Type + Horsepower, cars)
coef(ancova_mod)
summary(ancova_mod)$r.squared
summary(ancova_mod)$sigma
```

Interpretation of estimated coefficients? Why did $R^2$ increase? (We will be explaining this in detail in the next unit).

What's going on the design / model matrix?

```{r}
head(model.matrix(Price ~ Type + Horsepower, cars))
```

Same as model matrix with just `Type`. Since `Horsepower` is continuous, it doesn't get dummified to more features.

What if we went back to the `Type` regression, left out the intercept, dummified and added the intercept back in?

```{r}
Xmm = model.matrix(Price ~ 0 + Type, cars)
Xmm = cbind(1, Xmm)
head(Xmm,20)
```

Are the columns linearly independent? No ... so when we try to get the hat matrix,

```{r}
Xmm %*% solve(t(Xmm) %*% Xmm) %*% t(Xmm)
```

You can't invert a non-invertible matrix!!

What does R do when using the linear model function:

```{r}
coef(lm(cars$Price ~ 0 + Xmm))
```

SOMEHOW: it doesn't complain since it handles the non-invertibility (we don't know why...) but we do see that it's busted. Look at the coefficients! One is missing! What is it doing?? It's just arbitrarily dropping one (just like recommended).

