---
title: "Practice Lectures Week 1 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "Jan 25, 2024"
---

You can get started with this cheatsheet:
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

Acknowledgement: I would like to thank Professor Andreas Buja at Wharton as I cribbed lots of his notes. Here's the sources:
http://stat.wharton.upenn.edu/~buja/STAT-470-503-770/CHAPTERS/
http://stat.wharton.upenn.edu/~buja/STAT-961/

as well as Yonatan Rosenblatt previously of Ben Gurion University:
http://www.john-ros.com/Rcourse

## Basic R

What is R? 

* An open-source, free standard for statistical and numerical programming
* interpreted, scripted and high level (like Python and not like e.g. C++)
* Has a tremendous developer community and industry community
* I joke that I'm an R developer for a living since I publish research in statistical methodology which comes with R packages
* History: came from Bell Labs. C --> S --> R

Upsides? 
* Very good for "one-shot" scripting - very common in data science
* New stats algorithms and methodologies appear here first! Waaay before Python.

Downsides?
* Slow (but this can be fixed by inlining other languages). This is mostly due to the fact that it is an interpreted and dynamically-typed language. This means that the code is first "interpreted" and then converted to machine code and then executed (see https://subscription.packtpub.com/book/application_development/9781783989263/1/ch01lvl1sec09/r-is-interpreted-on-the-fly). This has advantages though.
* Bad threading / parallelization support (ditto)
* Bad support for hashmaps and other inconveniences

What is RStudio? 

* An integrated developer environment (IDE). It is not the only IDE for R, but it is the most popular at the time of this writing. It consists of the file window (what you're reading) and the console window (below) and a few other convenient windows which will be explained later.

File types:

* The standard R script has a .R extension
* This file if you note has an .Rmd extension meeaning "R Markdown". Markdown is a way of typestting simple documents with basic code directives. You can read up on this by yourself. "R Markdown" is a document type that combines R code within markdown documents where the R code is in "chunks".

Let's see some R! Since R is interpreted, you can run commands directly into its command line by using ctrl-enter line-by-line or ctrl-shift-enter for the entire "chunk":

```{r}
5
```

and

```{r}
7
8
9
```

These are not the most exciting R commands. The above is called a "chunk" and it tells the R Markdown compiler that there is R code inside those tick marks.

Here's some R code commented:

```{r}
5 #This will produce the number 5
```

Everything beyond the "#" is a comment as is ignored during interpretation.

Let's start by using R as a calculator

```{r}
5 + 5
5 - 5
5 / 5
5 * 5
5^5 #or 5 ** 5 (not as common)
5 / 3
100 * 1 + 0.08^5
100 * (1 + 0.08)^5 #note order of operations - use parentheses
5 %% 3
5 * 5 %% 3
(5 * 5) %% 3 #note order of operations - use parentheses
```

You can have multiple statements per line. The semicolon ends a statement like C++ but it is omitted usually because if the line ends, it is implied. 

```{r}
5 + 5; 5 - 5
```

Multiline statements are rarely used. Only do it to make a point to yourself and your teammates - code is written for you not for the computer.

You can also extend expressions for many lines if you have a dangling operator at the end of the line e.g. this plus sign:

```{r}
5 +
  5 +
  5
```
Avoid this unless the line gets so long it's difficult to read. Note the "+" at the end of the line so the interpreter knows to keep looking to the next line. It is conventional to use some form of indentation on lines past the first line so show it's one command split over lines.

Let's take a look at how to represent numbers using scientific notation

```{R}
3,125 #commas are illegal and note that RStudio finds basic errors for us
3.125e3 #scientific notation
3.125E3
```


Every language has a means of invoking functions. Here are some included famous math functions. R uses parentheses to invoke functions. The quantities within the parentheses are known as "parameters" or "arguments" to the function.

```{r}
sqrt(2)
log(2)
log(2, 2)
log(2, base = 2) #This is preferred: specify the argument by name after the first argument.
abs(-1)
floor(1.2)
ceiling(1.2)
round(1.2345)
round(1.2345, 2)
log10(1000)
sin(pi) #pi is a hard-wired numerical constant, the only numerical constant in the language
#note the numerical error
acos(1)
exp(1)
```

We can control the number of digits via

```{r}
options(digits = 3)
3.123456789E-6
options(digits = 7) #the default
3.123456789E-6
```


There are lots of options. Look at the help file for a function by using the question mark:

```{r}
?options
```

Let's talk about basic data types:

```{r}
TRUE #logical
FALSE
T #please don't use this constant in your programs; use TRUE.
F #please don't use this constant in your programs; use FALSE
class(TRUE)
1 #numeric, think of double type in C++ or Java
class(1)
1L
class(1L) #what's the L doing? It's forcing a certain type
```

There are more formal ways to "force" AKA "cast":

```{r}
class(as.integer(1))
class(as.numeric(1L)) #what's the L doing? It's forcing a certain type
1.L #what's this? Not an error but a warning
1e3 #scientific notation convenience
1E3
class(1E3)
"A"
class("A")
as.integer(TRUE)
as.numeric(TRUE)
as.integer(FALSE)
as.numeric(FALSE)
as.logical(0)
as.logical(1)
as.logical(0.1) #anything else besides zero...
as.character(1)
as.character(0.1)
as.logical("a") #more on this "character" later
as.logical("FALSE") #more on this "character" later
as.logical("T") #more on this "character" later
```

All of the above are actually 1-dim vectors. We will explore this later.

To create vectors, we use the "c" function which is short for "combine". This function takes any number of arguments

```{r}
c(1, 1, 2, 3, 5, 8)
```

We can also use shorthand "ladder" notation:

```{r}
1 : 3
1 : 100
-3 : 3
3 : -3
-(1 : 5)
```

The "seq" function (short for "sequence") is key for this too:

```{r}
seq(1, 9) #short for 1 : 9
seq(1, 9, by = 2)
seq(1, 9, length.out = 3)
seq(1, 9, length.out = 4)
```

and so is the "rep" function (short for "replicate"):

```{r}
rep(1, 10)
rep(c(1, 2, 3), 10)
rep(seq(1, 2, by = 0.1), 10)
```

Every language has a means to assign a piece of data to variables:

```{r}
x = 1 #I use this... only because I like using one less character and I like my code to read like other popular languages
x
x <- 1 #but this is actually preferred according to the the R spec
2 -> x #even those this is legal, please don't do this!!
x
T = "my string"
x_the_number = 1 #underscored variables preferred
xTheNumber = 1 #camel case is a different style
x.the.number = 1 #some people like this but I don't because it is confusing with other languages e.g. Java
```

And a means to see what variables have been assigned and remove these variables:

```{r}
ls() #who's out there
rm(x)
x
rm(x_the_number, xTheNumber, x.the.number)
rm(list = ls()) #clear the workspace completely!
```

Printing them out is important:

```{r}
print(x) #gives you the output numbered by the number of outputs
cat(x) #just prints and it is preferred
cat(x); cat(x) #it prints without a line break
cat(x, "\n") #you can have many arguments to cat and it will concatenate them
cat("x is", x, "\n"); cat("x is", x, "\n")
```
There is a means to use logical comparisons:

```{r}
x = 1
x == 1
x == 0
x != 1
x != 0
x > 1
x < 1
x >= 1
x <= 1
x = TRUE; y = FALSE
x & y #AND
x | y #OR
```

Let's talk about some simple operations on vectors:

```{r}
x = seq(1, 100)
length(x) #how long is this vector
head(x) #the first elements
tail(x) #the last elements
head(x, 11) #the first 11 elements
tail(x, 13) #the last 13 elements
x[1] #square brackets are the "getter" - it gets the value at that index. R has one-based indexing while C, C++, Python, Java, have zero-based indexing. Get used to it. SORRY!!!
x[1] = 1984 #square brackets also serve as the "setter" - it sets the value at that index
x[1]
x[6]
x[length(x)] #unfortunately no "-1" last index convenience...
x[c(2, 77)]
x[2 : 77]
x[seq(1, 100, by = 2)] #odd numbers
x[-1] #everything but 1
x[-(1 : 10)] #everything but 1 - 10
x[-seq(2, 100, by = 2)] #another way to see odd numbers
x = c(rep(1, 10), rep(2, 10), rep(3, 10))
unique(x) #only unique elements (in order in which they appear)
x[1 : 10] = 10 : 1 # you can set in vectors as well
x[1 : 10]

x = c(1:10, 3:40)
```

and comparisons on vectors

```{r}
x = 1 : 5
y = 1 : 5
x == y
all.equal(x, y)
```

Most operations in R are vector-operations and these are preferred because they are optimized:

```{r}
x = 1 : 5
x + 2
x * 2
x^2
```
A note about logical vectors:

```{r}
x = c(TRUE, TRUE)
y = c(TRUE, FALSE)
z = c(FALSE, FALSE)
!x
x & y
x | y
x && y #vector AND - confusing - recommend not to use this
x || y #vector OR - confusing - recommend not to use this
as.numeric(x)
sum(x) #does the numeric conversion for you
prod(x) #does the numeric conversion for you
any(x) #convenient function
any(z)
all(x)
all(y)
#sometimes a useful function
xor(TRUE, FALSE)
xor(TRUE, TRUE)
xor(FALSE, FALSE)
```

Sampling is very important:

```{r}
x = 1 : 100
sample(x, 3)
sample(x, 101)
sample(x, 3, replace = TRUE)
sort(sample(x, 101, replace = TRUE))
sample(x) #default is length of the vector i.e. just shuffles
```

There are many "illegal" computation values in R: NaN, Inf, -Inf,

```{r}
1 / 0 #unlike C, Java -- no error... handles this natively
1 / 0 == Inf #this works
0 / 0 #unlike C, Java -- no error... handles this natively
x = 0 / 0
x == NaN #beware!!
is.nan(x)
-1 / 0
1 + 1 / 0
1 + Inf
1 / Inf
1 / (-Inf)
1 / NaN
log(0)
log(-1)
sqrt(-1)
```

There are a couple more "illegalish" values that are non-computational: NA, NULL, empty vectors

```{r}
NA #this is the value of missing
x = NA
x == NA #beware!!
is.na(x)
x = NULL #special reserved primitive for data that is "undefined"
x == NULL #strange... and beware
is.null(x)
#let's look at that strange thing
x = 1 : 3
x[1]
x[0] #a vector with zero elements -- a "null vector" of sorts
length(x[0])
c() #remarkably... not what we expect...
x[NA]
x[NaN]
x[Inf]
x[4]
```

Now let's look at data types again. These types are the data types we learned in class.

```{r}
x = c(1, 2, 3, 4, 5)
class(x)
x = seq(1, 5)
class(x) #integer and numeric are approximately the same for purposes of this class
x = sample(rep(c(0, 1), 50))
x #binary variable?
x = c("a", "b", "c", "d")
x
class(x)
x = rep(x, 5)
x
class(x)
x = factor(x)
x
?factor
levels = c("low", "medium", "high")
x_char = sample(rep(levels, 10))
x = factor(x_char)
x #nominal categorical variable
x = factor(x_char, levels = levels)
x
x = factor(x_char, levels = levels, ordered = TRUE)
x #ordinal categorical variable
as.numeric(x) #coerce this variable to a number... result makes sense
```


Data which is of class factor can be used in mathematical models that we build just like numeric. It will, by default be reduced to binary variables.

## Realizations from r.v.'s

Let's now do some probability a la Math 241. Let's realize a few iid random variable models. Let's say $X_1, ..., X_{10} \iid$ ...

* binomial
* geometric
* poisson
* standard uniform --- all numbers between 0 and 1 equally likely
* exponential
* normal with mean 5 and standard deviation 2

```{r}
n = 10 #good programming practice to declare shared data

x = rbinom(n, size = 5, prob = 0.1)
x
dbinom(1, size = 5, p = 0.1)
pbinom(3, size = 5, p = 0.1)
qbinom(.75, size = 3, p = 0.1)
x #returns as a vector
?rbinom
x = rgeom(n, prob = 0.1)
x
?rgeom
x = rpois(n, lambda = 2)
x
?rpois
x = runif(n)
x
?runif
x = rexp(n, rate = 1)
x
?rexp
x = rnorm(n, mean = 5, sd = 2)
x
?rnorm
```

Randomness is an illustion in a computer. This is a whole topic in and of itself. But for our purposes, imagine random numbers being generated from a long but fixed list of numbers and the starting point is random. But you can make the randomness deterministic by setting the starting point. Then it's just the same numbers in the same order! This starting point is called a "seed" and it's an integer. 

```{r}
set.seed(1984)
rnorm(10)
set.seed(1984) #reset to that point in the list
rnorm(10) #same exact numbers!!!
set.seed(1984.1)
rnorm(10) #same - needs to be a different integer
set.seed(-1984)
rnorm(10) #different
set.seed(1983)
rnorm(10) #different
get.seed()
```

When writing programs that depend on randomness, you should set the seed at the top so the program is deterministic. Then it is easy to find bugs. When you switch the program to production, you just comment ou the set seed line. I will be doing this fairly often during demos for the rest of the semester.

Sorting is also a key skill in R:

```{r}
n = 50
x = rbinom(n, size = 20, prob = 0.2)
x
sort(x)
sort(x, decreasing = TRUE)
```

## Matrices

Now that we finished vectors, let's learn about the matrix object. Let's use the dimension of the learning problem in class.

```{r}
n = 100 #number of historical objects: the people
p = 3 #number of features about each

random_junk = round(runif(n * p), 2)
random_junk
X = matrix(random_junk, nrow = n, ncol = p)
X
class(X)
X[1, 1] #the square brackets are still the getter but since there's two dimensions there's two indices
X[1, ] #first row
class(X[1, ]) #note: not matrix anymore
X[1, , drop = FALSE]
class(X[1, , drop = FALSE]) #still matrix... this matters sometimes and we will see an example of it
X[, 1] #first column
class(X[, 1]) #note: not matrix anymore
X[, 1, drop = FALSE]
class(X[, 1, drop = FALSE])
#other ways of subsetting matrices
X[1 : 4, ]
X[, 2 : 3]
X[, c(1, 3)]
X[seq(1, n, by = 2), ]
X[1, 1] = 1984 #the squarae brackets are still the setter
X[1, 1] = "1984" #warning: this will wind up casting all values in the matrix to string
head(X)
```

Sometimes handy but not recommended is getting / setting with only one index.

```{r}
head(X)
X[1] #first entry i.e. the 1,1 entry
X[2] #first entry i.e. the 2,1 entry (column major indexing, see https://cran.r-project.org/web/packages/reticulate/vignettes/arrays.html)
X[n * p] #last entry i.e. n,p entry
X[n * p + 1] #beyond last entry returns missing / illegal
X[-1] #returns everything but the first entry and it automatically vectorizes
X[1 : 100] #returns the first 100 i.e. the first column
X[X > 0.5] = NA #replace all values no matter what row / col they exist in
X
```


A few handy functions for numerical matrices:

```{r}
X = matrix(random_junk, nrow = n, ncol = p) #reset
rowSums(X)
colSums(X)
rowMeans(X)
colMeans(X)
```

## Conditional statements and looping

We will now learn how to program in R.

* Conditional Statements (and ways to print to screen too).

```{r}
x = 3
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else {
  print("The value of x is"); print(x) #each gets its own line
}

if (x > 0){
  ##
} 
else { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!! This is not Java or C++!
  ##
}

x = -5
if (x > 0){
  cat("The value of x is", x, "\n") #concatenate and print to screen
} else if (x < -3) { ##NOTE: ELSE MUST GO ON SAME LINE AS THE IF'S CLOSE BRACE!!!
  print(paste("The value of x is", x, collapse = " ")) #paste with collapse is like a string join
} else {
  print("The value of x is"); print(x) #each gets its own line
}
```

There is a one line convenience function called ifelse:

```{r}
x = -10 : 10
b = -10 : 10
a = ifelse(x > 0, b, -1)
a
x = -5 : 5
a = ifelse(x > 0, 1, -1)
a
```



There is also a switch in R. It is _not_ an official keyword (despite RStudio's endorsement), but it is useful for non-vector character expressions:

```{r}
ch = "two"
# ch = c("one", "two")
switch(ch,
   "one" = 1,
   "two" = 2,
   "three" = 3,
   -99
)
```


* Looping

A for loop contains three ingredients: (a) a looping variable joined with keyword "in" to a (b) vector or list (which we will discuss later) and executes over an (c) expression recommended to be enclosed in curly braces:

```{r}
for (i in 1 : 10){
  my_loop_variable = i
  print(my_loop_variable)
}
```

A slight difference from most other functional languages: after the loop is executed both the looping variable (i) still exists and the variable(s) within the scope of the looped expression (my_loop_variable) still exist:

```{r}
i
my_loop_variable
```

It is recommended to "clean up your namespace" whenever you can.

```{r}
for (i in 1 : 10){
  my_loop_variable = i
  print(my_loop_variable)
}
rm(i, my_loop_variable)
```

Remember, these two variables would vanish in C++ / Java when the for loop ended as it goes out of scope.

You can iterate over a vector as well Python-style or Java-generic-style:

```{r}
vec = seq(1, 10)
for (i in vec){
  print(i)
}
```

And the vector does not have to be numeric:

```{r}
vec = c("a", "b", "c")
for (ch in vec){
  print(ch)
}
```

There is also the while loop but not a "do-while" loop

```{r}
x = 1
while (x <= 10){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
}
```

This can also be done using the while-true and a break:

```{r}
x = 1
while (TRUE){
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The "repeat" keyword is the same as the while-true. It is slightly more esoteric but still good to know:

```{r}
x = 1
repeat {
  print(x)
  x = x + 1 #note: there is no "++" incrementing in R
  if (x > 10){
    break
  }
}
```

The final control keyword tells the loop to break the current iteration but continue to the next:

```{r}
xs = rep(NA, 10)
xs[3] = 3
xs[7] = 7
xs

tot = 0
for (x in xs){
  if (is.na(x)){
    next #move to next element in the list in the for loop
  }
  tot = tot + x
}
tot
```

Note: "real" R programmers avoid loops since they are slow. Instead they use the "apply" function (but not "sapply" or "lapply" which are not faster). We will cover these in the future. The latest trend is to use in-line C++ code (we will get to this).

## More data types

Before we get back to modeling, it is worth knowing a couple more data structures in R. These are not "data [science] types", these are "[computer science] data types" that are used frequently in data science applications. 

The first are "lists" which are "ordered hashmaps" or "ordered dictionaries" or "hash tables". if You don't know what this is, you should read about this online as it should have been covered in a intro to CS class.

```{r}
dict = list()
dict$a = "first"
dict$b = "second"
dict$c = "third"
dict
length(dict)
names(dict) #keys
dict_unlisted = unlist(dict) #values
dict_unlisted
class(dict_unlisted) #i.e. a vector
#three ways to access values by key / ordered location
dict$a
dict[["a"]]
dict[[1]] #value of first entered key
class(dict[[1]])
dict[[1 : 2]] #bombs
#now let's try to access a value for a non-existent key / ordered location
dict$q
dict[["q"]]
dict[[4]] #bombs
#convenient means to subset the list
dict[1]
class(dict[1])
dict[1 : 2]
dict[1 : 4] #this is the reason this type of access is not recommended
dict = list("first", "second", "third") #no key => value... what happens?
dict #default keys are the numbers 1, 2, ...
dict[[1]]
dict = list("a" = "first", "b" = "second", "c" = "third") #key => value
dict
dict = list(a = "first", b = "second", c = "third") #key => value
dict
```

There is no "hash set" native to the language. You can create a dictionary and fill it with missings:


```{r}
dict = list()
dict$a = NA
dict$b = NA
dict$c = NA
dict
length(dict)
names(dict) #keys
```

Lists conveniently allow all sorts of data types (as values only).

```{r}
varied_dict = list()
varied_dict$a = "first"
varied_dict$b = 2
varied_dict$c = 1 : 7
varied_dict$d = matrix(NA, nrow = 2, ncol = 2)
varied_dict[["some function"]] = function(x){x^2} #this key is not recommended
varied_dict
varied_dict$`some function` #note the tick marks (sometimes seen) needed due to spaces in key name
length(varied_dict)
names(varied_dict)
```

They have lots of uses in data science applications. We will likely see them in class and if not, you'll definitely see them in the real world. Note that data.frame objects are implemented as lists as well as many other common R objects.

Unfortunately, list can only truly accept characters as keys. If you really need more flexibility here, we will need a library (coming soon).


We will now discuss arrays i.e. multidimensional vectors

```{r}
x = array(1 : 5, 5)
x
X = array(1 : 25, dim = c(5, 5))
X
X = array(1 : 125, dim = c(5, 5, 5))
X
X[1, , ]
X[, 1, ]
X[, , 1]
X[1, 1, 1]
```

These can be associative arrays too and operate like a hash of vectors across arbitrary dimensions:

```{r}
X = array(1 : 125, 
          dim = c(5, 5, 5),
          dimnames = list(
            c("A", "B", "C", "D", "E"),
            c("I", "II", "III", "IV", "V"),
            c("blue", "red", "green", "yellow", "orange")
          ))
X
X["A", , ]
X[, "III", ]
X[, , "orange"]
X["C", , "orange"]
X["C", "IV", "orange"]
```


* Functions

```{r}
my_function = function(x){
  x
}
##You may be used to:
# int my_function(int x){
#  //something
#  return x;
# }
```

* Functions are objects in R. This is why you are actually assigning the function to an object.
* We don't need to declare type since R is not "statically typed" (higher level languages usually are not statically typed). Objects can be coerced into different types on the fly (R is "dynamically typed").
* No need for a "return" statement as the last line is the data that is returned. It is considered bad style to use "return" in R.

Let's make sure this works:

```{r}
my_function(3)
my_function("asd")
my_function(x = 3) #you can specify that 3 is the value for argument "x"
my_function(y = 3) #illegal argument
(function(x){x + 1})(3) #anonymous function or "lambda" (see https://en.wikipedia.org/wiki/Anonymous_function)
```

R is somewhat user friendly as it allows for default argument values, making those arguments optional to specify when calling the function:

```{r}
my_function = function(x = 1, y = 2, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #default execution
my_function(p = 0) #one optional argument specified, others defaulted
my_function(y = -2, q = 0) #two optional arguments specified, others defaulted

my_function = function(x = 1, y, z = 3, p = 4, q = 5, r = 6){
  (x + y + z) / (p + q + r)
}
my_function() #no dice
my_function(1, 0) #required argument specified
my_function(y = 0, q = 7) #required argument specified and one optional argument
rm(my_function) #can be deleted since it's an object
```

There are also common functional programming functions. 

* Reduce uses a binary function to successively combine the elements of a given vector and a possibly given initial value. 
* Filter extracts the elements of a vector for which a predicate (logical) function gives true. 
* Find and Position give the first or last such element and its position in the vector, respectively. 
* Map applies a function to the corresponding elements of given vectors. 

If you like this, there are many packages that extend this and organize it nicely e.g. `purrr` (we will get to packages next class).

```{r}
x = c(1, 2, 3, 4, 5)
Reduce(sum, x)
Filter(function(x){x <= 3}, x)
Find(function(x){x > 4}, x)
unlist(Map(function(x){x + 100}, x)) #what happened here?? Map will return a list (for flexibility)
```

## Libraries in R

So far we've only made use of "base R". This is the funcionality included from a vanilla installation of R. 

R has a huge worldwide community of contributors. Some contribute to newer version of base R, but most create open-source "R packages" or "libraries". Many libraries come preinstalled. For instance, the MASS library which stands for "Modern Applied Statistics with S" (a famous textbook of R). We can call a function from the MASS library via the following:

```{r}
MASS::as.fractions(0.99)
MASS::as.fractions(pi)
```

Note we cannot just execute

```{r}
as.fractions(pi)
```

We made use of the scope operator "::" to access a namespace beyond the usual "global namespace" which we've been used to. Parenthetically, you can use the ":::" to access the private / internal functions and variables. Anyone who understands object-oriented programming with public interfaces / APIs would cringe at this!!!

If we are using the MASS library a lot, using the scope operator may get annoying. So similar to the "with" command, we can call

```{r}
library(MASS)
```

which loads all public methods (aka "exported" functions) into the public namespace. 

Now, after the library invocation we can do the following and treat it as a normal function:

```{r}
as.fractions(pi)
```

Is this always a good idea? No... everytime you call `library` it "dirties" the namespace by putting all the functions there and rewriting over functions there previously. This is bad because you are more likely to get namespace conflicts. For instance. Let's say package `kapelner` had a weird `sample` function. This would be clear:

```{r}
v = rnorm(100)
kapelner::sample(v)
sample(v)
```

The first line is doing the special sample function and the second is using base R's sample. But if I do this:

```{r}
library(kapelner)
#...
#...
###10,000 lines of code in which you forget about the fact that the kapelner library is loaded
#...
#...
sample(v)
```

You may think you're getting base R sample function, but you're not and now you're in  bug-city! You would have to do the following to be explicit:

```{r}
library(kapelner)
sample(v)
base::sample(v)
```

This is not a recommended thing to do. It's also not recommended for package developers to name functions the same as common base R functions. But this doesn't stop them!

Back to real packages... the content for the MASS package was sitting on the hard drive since it comes with R. But what if you want to use a package that does not come with R? We'll have to install the package just like pip for Python, Rubygems for Ruby, R has a package management system built in. For example, here's a useful package for time series / finance stuff:

```{r}
install.packages("tseries")
```

Note that it knew where to go online - it went to a CRAN mirror. CRAN is the official repository for R packages. Now that it's installed, step 2 is to load it into namespace so we can more seamlessly access its functionality.

```{r}
library(tseries)
```

That was a welcome message.

This library is really cool e.g.

```{r}
ibm_stock_history = get.hist.quote(instrument = "IBM", start = "2018-01-01", end = "2018-02-01")
ibm_stock_history
```

Is this a data frame?

```{r}
class(ibm_stock_history)
```

Nope - they made their own data type. If we had a unit on "writing your own R packages", I would explain how this is done but alas there is no time...

Let's say you're sharing your code with someone and one of your lines is loading a library e.g.

```{r}
library(a_library_my_computer_does_not_have_installed_yet)
```

And my computer doesn't have this library. Then we need to stop what we're doing and install. This could be annoying. Here is a convenience: use the pacman package that installs if necessary:

```{r}
if (!require("pacman")){install.packages("pacman")} #installs pacman if necessary but does not load it!
pacman::p_load(devtools) #ensures that devtools gets installed and loaded into the workspace but pacman does not (very tidy)!
```

It is typical to then have a few lines declaring all packages on top of your R/Rmd script file. Here is an example header from one of my projects. 

```{r}
#if (!require("pacman")){install.packages("pacman")}
#pacman::p_load(knitr, randomForest, dplyr, tidyverse, doParallel, xtable, pracma, yaml)
```

We will be seeing this in pretty much all our demos in the future. It is very rare to be coding in R without making use of packages beyond base R. I'm going to require the use of pacman for HW / projects, etc. It just makes code easier to share, run, etc.

The devtools package is important for modern R usage. It allows downloading R packages directly from source that are not even on CRAN. This allows you to get "bleeding edge" features. For example:

```{r}
install_github("yihui/knitr")
```

However this doesn't always work!

```{r}
install_github("hadley/ggplot2")
```

Why can this fail? Because the computer you're running this on is not setup for compiling C++. Admittedly, MAC's usually succeed here and Windows usually fails here. To make it succeed you need to install a separate program beyond R called Rtools. This is one of the big advantages of using Linux and MAC over Windows - Windows just is more buggy when it comes to "real coding" and it gets in the way when you're out there trying to get stuff done. Linux absolutely is the best here and because Linux is usually the production environment anyway, it may make sense to use it for all your assignments and coding anyway.

Note, you can use the pacman library for this type of installation too. So your header becomes:

```{r}
if (!require("pacman")){install.packages("pacman")}
pacman::p_load(devtools)
pacman::p_load_gh("hadley/ggplot2")
```


# Convenient Mapping Function for Lists with the purrr package

We first load the library.

```{r}
pacman::p_load(purrr)
```

We will see later that the library `purrr` is part of a collection of libraries called the `tidyverse`.

Now imagine you have a collection of objects in a list. For example, let's let the object be matrices with different sizes:

```{r}
my_matrix_list = list()
my_matrix_list[["first"]] = matrix(rnorm(9), nrow = 3)
my_matrix_list[["second"]] = matrix(rnorm(12), nrow = 2)
my_matrix_list[["third"]] = matrix(rnorm(8), nrow = 4)
my_matrix_list
```

And you want to operate on each of those objects and return a list. Let's say I want to get back the dimensions, or the first rows, or the average values and return the same keys:

```{r}
my_dims_list = modify(my_matrix_list, ~ dim(.x))
my_dims_list
my_first_rows_list = modify(my_matrix_list, ~ .x[1, ])
my_first_rows_list
my_avgs_list = modify(my_matrix_list, ~ mean(.x))
my_avgs_list
```

This is a very convenient function known as "mapping" in functional programming. It saves a few lines of code e.g. the first `modify` would be:

```{r}
my_dims_list = list() #make new list to store keys --> dimensions of original matrices
for (key in names(my_matrix_list)){ #iterate over all list by key
  .x = my_matrix_list[[key]] #get value at the key for this iteration
  my_dims_list[[key]] = dim(.x) #run function on value and save it to new list
}
my_dims_list
```

The above which takes 5 lines and is repeated again and again and again in code all takes one line using the `modify` function. 

The `modify` function uses funky syntax which is not standard R. And it doesn't have to be; packages are allowed to extend the language and use symbols to create their own little mini-language. The `.x` above is a dummy variable for the value in the iteration in the imagined for loop (like in my rewritten boilerplate code above). The "~" tilde symbol we will be seeing in base R later on in class but in a completely different context. Here it just means "run the following function".

Modify is just one of the functions in the `purrr` package. See the following cheatsheet for more convenient functions: https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf.


## Loading datasets from R packages

Since R is a language built for data and statistics, it has a ton of interesting data sets by default and even more that are contained in packages. There is really just one command to know:

```{r}
rm(list = ls())
data(iris) #load the iris dataset (as a data frame). This dataset is included in the package "datasets" which is autoloaded by default
class(iris)
?iris
#3 things I always do immediately upon getting a dataset
head(iris)
str(iris)
summary(iris)
```

Here is another very famous dataset

```{r}
MASS::Boston #this just references the object but does not load it into the environment
data(Boston) #error since package MASS is not loaded by default
data(Boston, package = "MASS") #package argument not needed if package loaded 
head(Boston)
```

Most data sets are names some descriptive name like "loandata" or "cars". R has so many datasets. Here they all are by package installed:

```{r}
data(package = .packages(all.available = TRUE))
```

