We first set directory of the package files:

```{r}
setwd("C:/Users/kapel/workspace/QC_MATH_342W_Spring_2024/labs")
```

We will now generate all the help files using the Roxygen code above each function:

```{r}
pacman::p_load(roxygen2)
roxygenise("QCbasementGBM", clean = TRUE)
```

To create the R package locally, We now run the following from terminal:

> R CMD INSTALL QCbasementGBM

Then we can load it like any other package from CRAN even though it sits locally on our machine.
We remember to set JAVA's RAM first since we are using YARF, an rJava package:

```{r}
options(java.parameters = "-Xmx8000m")
pacman::p_load(QCbasementGBM)
```

We can read the help files we wrote!

```{r}
?QCbasementGBM::gbm_fit
?QCbasementGBM::gbm_all_predictions
?QCbasementGBM::predict.qc_basement_gbm
```

We can actually use it to fit (see lab 11)

```{r}
set.seed(1)
n = 100
p = 3
X = matrix(rnorm(n * p), nrow = n)
bbeta = seq(-1, 1, length.out = p)
y = c(X %*% bbeta + rnorm(n))
y_binary = rbinom(n, 1, 1 / (1 + exp(-X %*% bbeta)))
X = data.frame(X)
```

Then we very simply build the model and predict:

```{r}
g_b = gbm_fit(X, y)
yhats = predict(g_b, X) #canonical R invocation!
head(yhats)
```

We can make sure this is the same as the lab:

```{r}
pacman::p_load(ggplot2)
ggplot(data.frame(y = y, yhat = predict(g_b, X))) + aes(x = y, y = yhat) + geom_point()
y_hats_by_m = gbm_all_predictions(g_b, X)
rmses_by_m = apply(y_hats_by_m, 2, function(y_hat){sqrt(mean((y - y_hat)^2))})
rmses_by_m
```

If we want to submit it to CRAN, we first build it as a compressed file via

> R CMD build QCbasementGBM

Then we check it for CRAN checks via

> R CMD check --as-cran QCbasementGBM

If it passes (it doesn't pass for a whole bunch of reasons), 
we can upload it to: https://cran.r-project.org/submit.html
