---
title: "Lab 6"
author: "Your Name Here"
output: pdf_document
---

#Polynomial Regression and Interaction Regression

We will work with the diamonds dataset again. Here we load up the dataset and convert all factors to nominal type:

```{r}
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
diamonds$cut =      factor(diamonds$cut, ordered = FALSE)
diamonds$color =    factor(diamonds$color, ordered = FALSE)
diamonds$clarity =  factor(diamonds$clarity, ordered = FALSE)
skimr::skim(diamonds)
```

Given the information above, what are the number of columns in the raw X matrix?

#TO-DO

Verify this using code:

```{r}
#TO-DO
```

Would it make sense to use polynomial expansions for the variables cut, color and clarity? Why or why not?

#TO-DO

Would it make sense to use log transformations for the variables cut, color and clarity? Why or why not?

#TO-DO

In order to ensure there is no time trend in the data, randomize the order of the diamond observations in D:.

```{r}
#TO-DO
```

Let's also concentrate only on diamonds with <= 2 carats to avoid the issue we saw with the maximum. So subset the dataset. Create a variable n equal to the number of remaining rows as this will be useful for later. Then plot it.

```{r}
#TO-DO
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```

Create a linear model of price ~ carat and gauge its in-sample performance using s_e.

```{r}
#TO-DO
```

Create a model of price ~ clarity and gauge its in-sample performance

```{r}
#TO-DO
```

Why is the model price ~ carat substantially more accurate than price ~ clarity?

#TO-DO

Create a new transformed feature ln_carat and plot it vs price.

```{r}
#TO-DO
ggplot(diamonds, aes(x = ln_carat, y = price)) + 
  geom_point()
```

Would price ~ ln_carat be a better fitting model than price ~ carat? Why or why not?

#TO-DO

Verify this by comparing R^2 and RMSE of the two models:

```{r}
#TO-DO
```

Create a new transformed feature ln_price and plot its estimated density:


```{r}
#TO-DO
ggplot(diamonds) + geom_histogram(aes(x = ln_price), binwidth = 0.01)
```


Now plot it vs carat.

```{r}
ggplot(diamonds, aes(x = carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ carat be a better fitting model than price ~ carat? Why or why not?

#TO-DO

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
#TO-DO
```

We just compared in-sample statistics to draw a conclusion on which model has better performance. But in-sample statistics can lie! Why is what we did valid?

#TO-DO

Plot ln_price vs ln_carat.

```{r}
ggplot(diamonds, aes(x = ln_carat, y = ln_price)) + 
  geom_point()
```

Would ln_price ~ ln_carat be the best fitting model than the previous three we considered? Why or why not?

Verify this by computing s_e of this new model. Make sure these metrics can be compared apples-to-apples with the previous.

```{r}
#TO-DO
```

Compute b, the OLS slope coefficients for this new model of ln_price ~ ln_carat.

```{r}
#Model A
#TO-DO
```

Interpret b_1, the estimated slope of ln_carat.

#TO-DO

Interpret b_0, the estimated intercept.

#TO-DO

Create other features ln_x, ln_y, ln_z, ln_depth, ln_table.

```{r}
#TO-DO
```

From now on, we will be modeling ln_price (not raw price) as the prediction target. 

Create a model (B) of ln_price on ln_carat interacted with clarity and compare its performance with the model (A) ln_price ~ ln_carat.

```{r}
#Model B
#TO-DO
```

Which model does better? Why?

#TO-DO

Create a model of (C) ln_price on ln_carat interacted with every categorical feature (clarity, cut and color) and compare its performance with model (B)

```{r}
#Model C
#TO-DO
```

Which model does better? Why?

#TO-DO

Create a model (D) of ln_price on every continuous feature (logs of carat, x, y, z, depth, table) interacted with every categorical feature (clarity, cut and color) and compare its performance with model (C).

```{r}
#Model D
#TO-DO
```

Which model does better? Why?

#TO-DO

What is the p of this model D? Compute with code.

```{r}
#TO-DO
```

Create model (E) which is the same as before except create include the raw features interacted with the categorical features and gauge the performance against (D).

```{r}
#Model E
#TO-DO
```

Which model does better? Why?

#TO-DO

Create model (F) which is the same as before except also include also third degree polynomials of the continuous features interacted with the categorical features and gauge performance against (E). By this time you're getting good with R's formula syntax!

```{r}
#Model F
#TO-DO
```

Which model does better? Why?

#TO-DO

Can you think of any other way to expand the candidate set curlyH? Discuss.

#TO-DO

We should probably assess oos performance now. Sample 2,000 diamonds and use these to create a training set of 1,800 random diamonds and a test set of 200 random diamonds. Define K and do this splitting:

```{r}
#TO-DO
```

Compute in and out of sample performance for models A-F. Use s_e as the metric (standard error of the residuals). Create a list with keys A, B, ..., F to store these metrics. Remember the performances here will be worse than before since before you're using nearly 52,000 diamonds to build a model and now it's only 1,800! 

```{r}
#TO-DO
```

You computed oos metrics only on n_* = 200 diamonds. What problem(s) do you expect in these oos metrics?

#TO-DO

To do the K-fold cross validation we need to get the splits right and crossing is hard. I've developed code for this already. Run this code.

```{r}
temp = rnorm(n)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
head(folds_vec, 200)
```

Comment on what it does and how to use it to do a K-fold CV:

#TO-DO

Do the K-fold cross validation for model F and compute the overall s_e and s_s_e. 

```{r}
#TO-DO
```

Does K-fold CV help reduce variance in the oos s_e? Discuss.

#TO-DO

Imagine using the entire rest of the dataset besides the 2,000 training observations divvied up into slices of 200. Measure the oos error for each slice on Model F in a vector `s_e_s_F` and compute the `s_s_e_F` and also plot it.

```{r}
#TO-DO
ggplot(data.frame(s_e_s_F = s_e_s_F)) + geom_histogram(aes(x = s_e_s_F))
```



#Visualization with the package ggplot2

I highly recommend using the [ggplot cheat sheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) as a reference resource. You will see questions that say "Create the best-looking plot". Among other things you may choose to do, remember to label the axes using real English, provide a title and subtitle. You may want to pick a theme and color scheme that you like and keep that constant throughout this lab. The default is fine if you are running short of time.

Load up the `GSSvocab` dataset in package `carData` as `X` and drop all observations with missing measurements. This will be a very hard visualization exercise since there is not a good model for vocab.

```{r}
#TO-DO
```

Briefly summarize the documentation on this dataset. What is the data type of each variable? What do you think is the response variable the collectors of this data had in mind?

#TO-DO

Create two different plots and identify the best-looking plot you can to examine the `age` variable. Save the best looking plot as an appropriately-named PDF.

```{r}
#TO-DO
```

Create two different plots and identify the best looking plot you can to examine the `vocab` variable. Save the best looking plot as an appropriately-named PDF.

```{r}
#TO-DO
```

Create the best-looking plot you can to examine the `ageGroup` variable by `gender`. Does there appear to be an association? There are many ways to do this.

```{r}
#TO-DO
```

Create the best-looking plot you can to examine the `vocab` variable by `age`. Does there appear to be an association?

```{r}
#TO-DO
```

Add an estimate of $f(x)$ using the smoothing geometry to the previous plot. Does there appear to be an association now?

```{r}
#TO-DO
```

Using the plot from the previous question, create the best looking plot overloading with variable `gender`. Does there appear to be an interaction of `gender` and `age`?

```{r}
#TO-DO
```


Using the plot from the previous question, create the best looking plot overloading with variable `nativeBorn`. Does there appear to be an interaction of `nativeBorn` and `age`?

```{r}
#TO-DO
```

Create two different plots and identify the best-looking plot you can to examine the `vocab` variable by `educGroup`. Does there appear to be an association?

```{r}
#TO-DO
```

Using the best-looking plot from the previous question, create the best looking overloading with variable `gender`. Does there appear to be an interaction of `gender` and `educGroup`?

```{r}
#TO-DO
```

Using facets, examine the relationship between `vocab` and `ageGroup`. You can drop year level `(Other)`. Are we getting dumber?

```{r}
#TO-DO
```
