---
title: "Lab 9"
author: "Your Name Here"
output: pdf_document
---

#YARF

For the next couple of labs, I want you to make some use of a package I wrote that offers convenient and flexible tree-building and random forest-building. Make sure you have a JDK installed first

https://www.oracle.com/java/technologies/downloads/

Then try to install rJava

```{r}
options(java.parameters = "-Xmx4000m")
pacman::p_load(rJava)
.jinit()
```

If you have error, messages, try to google them. Everyone has trouble with rJava!

If that worked, please try to run the following which will install YARF from my github:

```{r}
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)
```

Please try to fix the error messages (if they exist) as best as you can. I can help on slack.

#Bagged Trees and Random Forest


Take a sample of n = 2000 observations from the diamonds data.

```{r}
#TO-DO
```

Find the bootstrap s_e for a RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can calculate oob residuals via `e_oob = y_train - rf_mod$predicted`. Plot.

```{r}
#TO-DO
```

Using the diamonds data, find the oob s_e for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. If you are using the `randomForest` package, you can create the bagged tree model via setting an argument within the RF constructor function. Plot.

```{r}
#TO-DO
```


What is the percentage gain / loss in performance of the RF model vs bagged trees model?

```{r}
#TO-DO
```

Why was this the result?

#TODO

Plot oob s_e by number of trees for both RF and bagged trees.

```{r}
#TO-DO
```

Build RF models for 500 trees using different `mtry` values: 1, 2, ... the maximum. That maximum will be the number of features assuming that we do not binarize categorical features if you are using `randomForest` or the number of features assuming binarization of the categorical features if you are using `YARF`. Calculate oob s_e for all mtry values. Plot.

```{r}
#TO-DO
```

Plot oob s_e by mtry.

```{r}
#TO-DO
```

```{r}
rm(list = ls())
```


Take a sample of n = 2000 observations from the adult data.

```{r}
#TO-DO
```

Using the adult data, find the bootstrap misclassification error for an RF model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees.

```{r}
#TO-DO
```

Using the adult data, find the bootstrap misclassification error for a bagged-tree model using 1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500, 1000 trees. Plot.

```{r}
#TO-DO
```

What is the percentage gain / loss in performance of the RF model vs bagged trees model?

```{r}
#TO-DO
```

Plot bootstrap misclassification error by number of trees for both RF and bagged trees.

```{r}
#TO-DO
```

Build RF models for 500 trees using different `mtry` values: 1, 2, ... the maximum (see above as maximum is defined by the specific RF algorithm implementation). Plot.

```{r}
#TO-DO
```

Plot bootstrap misclassification error by `mtry`.

```{r}
#TO-DO
```